---
title: "Clustering embeddings with EVoC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering embeddings with EVoC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

Embedding models turn text, images and other objects into vectors whose
directions carry meaning: two posts about the same thing point the same way.
Clustering such vectors is how topics are discovered in a corpus, and it is
what `shoal_evoc()` is for. This vignette runs it on real sentence embeddings,
looks at what it returns, and compares it with the other ways of clustering
the same vectors.

```{r setup, message = FALSE}
library(shoal)
```

## The data

`newsgroups` holds 2,400 posts from eight groups of the 20 Newsgroups corpus,
embedded with a sentence-transformer and reduced to 64 dimensions (see
`?newsgroups` for how). The group each post came from is kept for scoring,
along with the opening of each post for reading.

```{r data}
str(newsgroups)
table(newsgroups$group)
```

Scores below are the adjusted Rand index (ARI) against the group labels: 1 for
a perfect match, 0 for chance. Noise counts as its own class.

```{r ari}
ari <- function(a, b) {
  a <- ifelse(is.na(a), -1L, as.integer(factor(a)))
  b <- ifelse(is.na(b), -1L, as.integer(factor(b)))
  tab <- table(a, b)
  comb2 <- function(x) x * (x - 1) / 2
  sum_ij <- sum(comb2(tab))
  sum_a <- sum(comb2(rowSums(tab)))
  sum_b <- sum(comb2(colSums(tab)))
  expected <- sum_a * sum_b / comb2(sum(tab))
  (sum_ij - expected) / ((sum_a + sum_b) / 2 - expected)
}
```

## One call

EVoC needs no number of clusters and no distance threshold. It builds a
nearest-neighbour graph under cosine distance, learns a compact node
embedding of that graph, and density-clusters the embedding at every
granularity it can find. `min_cluster_size` is the one parameter worth
setting up front: the upstream default of 5 suits corpora of hundreds of
thousands of documents, and over-fragments a few thousand.

```{r fit}
x <- newsgroups$embedding
timing <- system.time(ev <- shoal_evoc(x, min_cluster_size = 30L))
ev
timing[["elapsed"]]
```

```{r fit-table}
table(cluster = ev$cluster, group = newsgroups$group, useNA = "ifany")
ari(newsgroups$group, ev$cluster)
```

Seven of the eight groups come back as a cluster each, with a tenth of the
corpus set aside as noise. The eighth, `talk.politics.mideast`, comes back as
two clusters. The labels say that is an error; the posts say otherwise. The
opening lines of a few from each show that the group holds two separate
arguments, one about Turkey, Armenia and Greece and one about Israel and the
Holocaust, and EVoC has found a distinction the labels do not have.

```{r split}
mideast <- table(ev$cluster[newsgroups$group == "talk.politics.mideast"])
halves <- as.integer(names(sort(mideast, decreasing = TRUE)[1:2]))

set.seed(2)
for (h in halves) {
  members <- which(!is.na(ev$cluster) & ev$cluster == h)
  cat("Cluster", h, "\n")
  cat(paste0("  ", substr(newsgroups$snippet[sample(members, 4)], 1, 90)), sep = "\n")
}
```

## Every layer, not one

A corpus rarely has one right granularity. EVoC therefore returns every
stable layer of the cluster hierarchy, finest first, and only chooses which
one fills `cluster`: by default the most persistent, matching the reference
implementation. The others are on the result.

```{r layers}
layer_summary <- data.frame(
  layer = seq_along(ev$layers),
  clusters = vapply(ev$layers, function(l) length(unique(l[!is.na(l)])), integer(1)),
  noise = vapply(ev$layers, function(l) sum(is.na(l)), integer(1)),
  persistence = round(ev$persistence, 1),
  ari = round(vapply(ev$layers, function(l) ari(newsgroups$group, l), numeric(1)), 3)
)
layer_summary
```

Choosing a different layer afterwards is an index, not a refit:

```{r layer-pick}
finest <- ev$layers[[1]]
table(finest, newsgroups$group, useNA = "ifany")[1:6, ]
```

## Seeing the clusters

The node embedding EVoC learned is on the result too. It has four dimensions
here, and the clustering used all of them, so the first two give a rough
picture rather than an exact one: clusters that overlap in this view are
separated in the other two.

```{r embedding-plot}
k <- ev$n_clusters
plot(ev$embedding[, 1:2],
     col = ifelse(is.na(ev$cluster), "grey60", shoal_palette(k)[ev$cluster]),
     pch = ifelse(is.na(ev$cluster), 4, 19), cex = 0.5,
     xlab = "Embedding 1", ylab = "Embedding 2",
     main = "EVoC node embedding, coloured by cluster")
```

Each observation also has a membership strength for its cluster in
`strengths`: 1 for a point that joined its cluster at full density, lower
for one that only attached as the density threshold fell. Two thirds of
members score exactly 1, so the strength is a flag for the doubtful rather
than a ranking of the core. The weakest members of the `rec.autos` cluster
are a mixed bag: a post from another group, and posts only loosely about
cars.

```{r strengths}
s <- ev$strengths[[ev$layer]]
mean(s[!is.na(ev$cluster)] < 1)

autos <- as.integer(names(which.max(table(ev$cluster[newsgroups$group == "rec.autos"]))))
members <- which(!is.na(ev$cluster) & ev$cluster == autos)
weakest <- members[order(s[members])][1:5]
data.frame(strength = round(s[weakest], 2), group = newsgroups$group[weakest],
           snippet = substr(newsgroups$snippet[weakest], 1, 55))
```

## Against the alternatives

Three other ways to cluster the same vectors, with what each needs from you.

```{r alternatives, message = FALSE}
library(uwot)

score <- function(name, cluster, seconds) {
  data.frame(method = name,
             clusters = length(unique(cluster[!is.na(cluster)])),
             noise = sum(is.na(cluster)),
             ari = round(ari(newsgroups$group, cluster), 3),
             seconds = round(seconds, 2))
}

# HDBSCAN straight on the vectors, with the cosine metric.
t_hdb <- system.time(
  hdb <- shoal_hdbscan(x, min_cluster_size = 15L, min_samples = 10L,
                       metric = "cosine", boruvka = FALSE)
)

# UMAP to two dimensions under cosine distance, then HDBSCAN.
set.seed(42)
t_umap <- system.time({
  u <- umap(x, n_neighbors = 30, min_dist = 0, metric = "cosine", n_threads = 2)
  uh <- shoal_hdbscan(u, min_cluster_size = 40L, min_samples = 10L)
})

# k-means, told the right number of groups.
t_km <- system.time(km <- shoal_kmeans(x, k = 8L))

rbind(
  score("EVoC", ev$cluster, timing[["elapsed"]]),
  score("HDBSCAN on the vectors", hdb$cluster, t_hdb[["elapsed"]]),
  score("UMAP then HDBSCAN", uh$cluster, t_umap[["elapsed"]]),
  score("k-means, k = 8", km$cluster, t_km[["elapsed"]])
)
```

HDBSCAN on the raw vectors is the case EVoC exists to fix: in 64 dimensions
there is no density contrast to find, so most of the corpus is noise, and the
spatial index is slow getting there. UMAP followed by HDBSCAN is the usual
remedy and works, at the cost of two stages, two sets of parameters, a
stochastic layout, and tens of times the running time. EVoC does the same
job in one call, and here does it a little better.

k-means scores well when told there are eight groups. It always will on a
corpus with round, evenly sized topics, and it has no way of saying when
that is not so: it cannot report noise, cannot find eleven topics when asked
for eight, and offers no layers. With a known `k` and a clean corpus it is
hard to beat; discovering the structure of an unknown one is the case for
EVoC.

## Stability

EVoC is stochastic, so `seed` matters, and as with any such method the
question is whether another seed finds the same clusters.

```{r seeds}
runs <- lapply(1:3, function(seed) shoal_evoc(x, min_cluster_size = 30L, seed = seed)$cluster)
round(c(seeds_1_2 = ari(runs[[1]], runs[[2]]),
        seeds_1_3 = ari(runs[[1]], runs[[3]]),
        seeds_2_3 = ari(runs[[2]], runs[[3]])), 3)
```

The runs agree on most of the partition. Where they differ is at the edges:
which doubtful posts are noise, and whether a split like the one in
`talk.politics.mideast` is made. On a corpus of a few thousand posts that
is the expected amount of variation; it shrinks as the corpus grows, which
is the regime EVoC is built for. The same seed, data and parameters always
give bitwise-identical results, whatever the thread count.

## Practical notes

- Input must be embedding vectors: rows whose direction carries the meaning.
  EVoC normalises them and works in cosine geometry throughout. Tabular
  measurements are not that; use the other algorithms, and see
  `vignette("umap")` for wide tabular data.
- Raise `min_cluster_size` first. It sets the finest granularity, and the
  default of 5 is calibrated for corpora far larger than most.
- Read the layers before trusting the selected one. Persistence is a good
  heuristic, not a law; on nested topic structure the finest layer is often
  the useful one.
- Expect some noise. Posts that are short, off-topic or between topics land
  in no cluster, which is information rather than failure.
