---
title: "Get started with shoal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with shoal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

shoal provides six clustering algorithms behind one interface. Each takes a
numeric matrix or data frame, returns an object with the shared
`shoal_clustering` class, and prints and plots the same way. Noise points, where
an algorithm has them, are `NA` in the `cluster` vector.

This vignette introduces each algorithm in turn: what it assumes, the
parameters that matter, and what its clusters look like on data chosen to show
its character. It then turns to the tools around them: the distance matrix
and nearest-neighbour search the algorithms rest on, and the indices for
choosing the number of clusters.

```{r setup}
library(shoal)
```

## What every result has in common

```{r shared}
x <- as.matrix(iris[, 1:4])
fit <- shoal_kmeans(x, k = 3L)

fit$cluster[1:10]
fit$n_clusters
fit$n_noise
fit$params
```

`data` holds the matrix the model was fitted to, so `plot()` needs nothing
else. With two columns it draws a scatter plot; with more, a scatter plot
matrix, or a chosen pair via `xcol` and `ycol`. Colours come from
`shoal_palette()`, and `col` and `pch` can be given directly to colour points
by something other than their cluster, as some of the figures below do.

## k-means: `shoal_kmeans()`

Partitions the data into `k` compact clusters of similar spread by minimising
the within-cluster sum of squares. It is fast and predictable, and the right
baseline when clusters are roughly spherical and you know how many to expect.
It is stochastic in its initialisation; `seed` controls that, and R's own
random number generator has no effect on it.

```{r kmeans}
km <- shoal_kmeans(x, k = 3L, n_runs = 10L, seed = 1L)
km
```

The centroids are the model. Overlaying them shows what k-means actually
found: three centres, with every point assigned to the nearest one.

```{r kmeans-plot}
plot(km, xcol = "Petal.Length", ycol = "Petal.Width")
points(km$centroids[, "Petal.Length"], km$centroids[, "Petal.Width"],
       pch = 4, cex = 2.5, lwd = 3)
```

Because it has cluster centres, it can assign new observations.

```{r kmeans-predict}
predict(km, x[c(1, 51, 101), ])
```

## Gaussian mixture: `shoal_gmm()`

Fits `k` multivariate Gaussians by expectation-maximisation. Each component
has its own covariance matrix, so clusters can be elongated and tilted, and
correlated or differently scaled features are handled without pre-scaling.
Every observation gets a probability of belonging to each component;
`cluster` is the most likely one and `posterior` holds the full matrix.

The difference from k-means is easiest to see on clusters that are not round.
Two long, parallel ellipses and a round blob:

```{r gmm-data}
ellipse <- function(n, centre, angle, sx, sy) {
  pts <- cbind(rnorm(n, sd = sx), rnorm(n, sd = sy))
  rot <- matrix(c(cos(angle), sin(angle), -sin(angle), cos(angle)), 2)
  sweep(pts %*% rot, 2, centre, "+")
}

set.seed(3)
angle <- pi / 6
ell <- rbind(
  ellipse(250, c(0, 0), angle, 2.5, 0.4),
  ellipse(250, c(-3.5 * sin(angle), 3.5 * cos(angle)), angle, 2.5, 0.4),
  ellipse(150, c(6, -3), 0, 0.7, 0.7)
)
colnames(ell) <- c("x", "y")

gm <- shoal_gmm(ell, k = 3L)
gm
```

k-means cuts across the two ellipses, because its boundaries are always
midway between centres. The mixture follows their shape. In the right-hand
panel, points are drawn fainter the less certain their assignment, using the
posterior and the `col` argument.

```{r gmm-plot, fig.height = 4}
par(mfrow = c(1, 2))
km_ell <- shoal_kmeans(ell, k = 3L)
plot(km_ell, col = shoal_palette(3)[km_ell$cluster], pch = 19)

confidence <- apply(gm$posterior, 1, max)
base <- col2rgb(shoal_palette(3)[gm$cluster]) / 255
alpha <- pmin(1, 0.15 + 0.85 * (confidence - 1 / 3) / (2 / 3))
plot(gm, col = rgb(base[1, ], base[2, ], base[3, ], alpha), pch = 19)
```

A fitted mixture has a `logLik()` method, so `AIC()` and `BIC()` work on it
directly; see [Choosing the number of clusters](#choosing-k). Only full
covariance matrices are supported.

## DBSCAN: `shoal_dbscan()`

Grows clusters from points that have at least `min_samples` neighbours within
radius `eps`, and labels everything else noise. Clusters can be any shape,
which no centroid-based method can offer. The bundled `rings` data, three
concentric rings plus scattered noise, is the standard case.

```{r dbscan}
db <- shoal_dbscan(rings, eps = 0.2, min_samples = 5L)
db
plot(db)
```

`eps` is on the scale of the data, so a value never carries between
datasets. `metric = "cosine"` is available for directional data.

## HDBSCAN: `shoal_hdbscan()`

DBSCAN's limitation is that `eps` is one global threshold, so every cluster
has to have the same density. HDBSCAN removes it by building the hierarchy of
clusterings across every density at once and keeping the most stable ones.
There is no `eps`; `min_cluster_size` sets the smallest group worth calling a
cluster and `min_samples` how conservative the density estimate is.

Two tight blobs beside one large diffuse cloud, with uniform noise around
them, shows the difference. No single `eps` serves both densities: small
enough for the blobs, it keeps only the core of the cloud and calls the rest
noise; large enough for the cloud, everything merges into one cluster.

```{r hdbscan-data}
set.seed(4)
vd <- rbind(
  cbind(rnorm(150, 0.0, 0.15), rnorm(150, 0, 0.15)),
  cbind(rnorm(150, 1.2, 0.15), rnorm(150, 0, 0.15)),
  cbind(rnorm(400, 0.5, 1.0), rnorm(400, 4, 1.0)),
  cbind(runif(60, -3, 4), runif(60, -2, 7))
)
colnames(vd) <- c("x", "y")

hdb <- shoal_hdbscan(vd, min_cluster_size = 30L, min_samples = 10L)
hdb
```

```{r hdbscan-plot, fig.width = 8, fig.height = 4.2}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot(shoal_dbscan(vd, eps = 0.3, min_samples = 10L), main = "DBSCAN, eps = 0.3")
plot(shoal_dbscan(vd, eps = 0.8, min_samples = 10L), main = "DBSCAN, eps = 0.8")
plot(hdb, main = "HDBSCAN")
```

HDBSCAN also returns a GLOSH outlier score for every observation, near 0 in
the core of a cluster and near 1 far from any. Drawn as a shade rather than a
cluster colour, it grades the whole dataset by how much it belongs.

```{r hdbscan-glosh}
plot(hdb, col = grey(0.85 * (1 - hdb$outlier_scores)), pch = 19,
     main = "GLOSH outlier score (darker = more outlying)")
```

Both density algorithms rely on a spatial index that degrades as the number of
columns grows. For wide data, either reduce the dimension first (see the UMAP
vignette) or, for embedding vectors, use EVoC.

## Hierarchical clustering: `shoal_hclust()`

Merges observations bottom-up into a dendrogram, which can then be cut at any
number of clusters without refitting. It takes a `dist` object, which
`shoal_dist()` computes in Rust, or raw data. The result is a standard
`hclust`, so `cutree()`, `plot()`, `rect.hclust()` and `as.dendrogram()` all
apply.

```{r hclust}
d <- shoal_dist(x)
hc <- shoal_hclust(d, method = "ward")
hc

groups <- cutree(hc, k = 3)
table(groups, iris$Species)
```

```{r hclust-plot, fig.height = 4}
par(mfrow = c(1, 2))
plot(hc, labels = FALSE, hang = -1, main = "Ward linkage", xlab = "", sub = "")
rect.hclust(hc, k = 3, border = shoal_palette(3))
plot(x[, "Petal.Length"], x[, "Petal.Width"], col = shoal_palette(3)[groups],
     pch = 19, xlab = "Petal.Length", ylab = "Petal.Width", main = "Cut at k = 3")
```

Seven linkage methods are available. Note that `"ward"` is R's `"ward.D2"`,
and that `"centroid"` and `"median"` take plain distances where
`stats::hclust()` expects squared ones; see `?shoal_hclust`.

## EVoC: `shoal_evoc()`

For embedding vectors: rows that are directions in a high-dimensional space,
such as the output of a text or image model. EVoC builds a nearest-neighbour
graph under cosine distance, learns a compact node embedding from it, and
density-clusters that embedding at several granularities at once. All layers
are returned; `layer` picks which one populates `cluster`, by default the most
persistent.

Eight topics of unequal size as directions in 48 dimensions, plus 100
scattered points:

```{r evoc}
set.seed(1)
sizes <- c(400, 300, 250, 200, 150, 100, 60, 40)
centres <- matrix(runif(length(sizes) * 48, -1, 1) * 0.6, nrow = length(sizes))
emb <- rbind(
  centres[rep(seq_along(sizes), times = sizes), ] +
    matrix(rnorm(sum(sizes) * 48, sd = 0.1), ncol = 48),
  matrix(runif(100 * 48, -1, 1) * 0.6, ncol = 48)
)

ev <- shoal_evoc(emb, min_cluster_size = 15L)
ev
```

The clusters cannot be drawn in 48 dimensions, but the node embedding EVoC
learned on the way is on the result. Its first two dimensions show every
layer: the finest one, fragmented and full of noise, and the persistent one
that was selected.

```{r evoc-plot, fig.height = 4}
draw_layer <- function(i) {
  layer <- ev$layers[[i]]
  k <- length(unique(layer[!is.na(layer)]))
  plot(ev$embedding[, 1:2],
       col = ifelse(is.na(layer), "grey60", shoal_palette(k)[layer]),
       pch = ifelse(is.na(layer), 4, 19), cex = 0.6,
       xlab = "Embedding 1", ylab = "Embedding 2",
       main = sprintf("Layer %d: %d clusters", i, k))
}
par(mfrow = c(1, 2))
draw_layer(1)
draw_layer(ev$layer)
```

EVoC is the wrong tool for ordinary tabular data: it assumes cosine geometry
and is calibrated for thousands to millions of rows. The upstream default
`min_cluster_size = 5` over-fragments small collections, so raise it first.

## Distances and neighbours: `shoal_dist()` and `shoal_knn()`

Both take the same nine metrics: the six of `stats::dist()`, cosine,
correlation and Mahalanobis. `shoal_dist()` returns a plain `dist`, so it
drops into `cmdscale()`, `cluster::pam()` or anything else that wants one.
`shoal_knn()` keeps only the `k` nearest neighbours of each row, so where a
distance matrix stops being possible somewhere in the tens of thousands of
rows, a neighbour search does not. It is exact, by a kd-tree in a few
dimensions and a parallel scan beyond that, and the two give identical
results. `vignette("distances")` goes through the choice of metric and what
a neighbour result can be turned into.

```{r knn}
nn <- shoal_knn(rings, k = 4L)
nn
head(nn$id, 3)
```

Its `plot()` is the standard way to choose `eps` for DBSCAN: each point's
distance to its `k`-th neighbour, sorted. Points on a ring have a small,
steady value; the noise points sit on the sharp rise at the right. `eps`
belongs at the elbow, which here is the 0.2 used above, with
`min_samples = k + 1` because DBSCAN counts the point itself.

```{r knn-plot, fig.height = 4}
plot(nn)
abline(h = 0.2, lty = 2)
```

## Choosing the number of clusters {#choosing-k}

`shoal_metrics()` computes the Calinski-Harabasz index (higher is better) and
Davies-Bouldin index (lower is better) for any partition, from the data.
`shoal_silhouette()` scores every observation from a distance matrix, and
works for any cluster shape. A fitted mixture's `BIC()` penalises parameters,
so it has an interior optimum where k-means' inertia does not.

```{r choose-k}
ks <- 2:8
indices <- do.call(rbind, lapply(ks, function(k) {
  shoal_metrics(shoal_kmeans(x, k = k))
}))
indices$bic <- vapply(ks, function(k) BIC(shoal_gmm(x, k = k)), numeric(1))
indices$silhouette <- vapply(ks, function(k) {
  attr(shoal_silhouette(d, shoal_kmeans(x, k = k)), "avg_width")
}, numeric(1))
indices
```

```{r choose-k-plot, fig.width = 8, fig.height = 2.8}
par(mfrow = c(1, 4), mar = c(4, 4, 3, 1))
for (m in c("calinski_harabasz", "davies_bouldin", "bic", "silhouette")) {
  plot(ks, indices[[m]], type = "b", pch = 19, xlab = "k", ylab = "", main = m)
}
```

Two clusters or three: the indices disagree, and that is the right answer,
because two of the species overlap. shoal reports the evidence and leaves the
decision to you.

## Which algorithm?

| You have                                       | Start with          |
|------------------------------------------------|---------------------|
| Compact clusters, a known `k`                  | `shoal_kmeans()`    |
| Elliptical clusters, or you want probabilities | `shoal_gmm()`       |
| Arbitrary shapes at one density                | `shoal_dbscan()`    |
| Arbitrary shapes at varying density            | `shoal_hdbscan()`   |
| A dendrogram, or cuts at several `k`           | `shoal_hclust()`    |
| Embedding vectors, thousands of rows or more   | `shoal_evoc()`      |
