---
title: "Get started with shoal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with shoal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

shoal provides six clustering algorithms behind one interface. Each takes a
numeric matrix or data frame, returns an object with the shared
`shoal_clustering` class, and prints and plots the same way. Noise points, where
an algorithm has them, are `NA` in the `cluster` vector.

This vignette introduces each algorithm in turn: what it assumes, the
parameters that matter, and what it returns beyond the cluster vector. Every
example uses the same two datasets, so the algorithms can be compared directly.

```{r setup}
library(shoal)

# Four measurements of 150 irises: three species, two of which overlap.
x <- as.matrix(iris[, 1:4])

# Three concentric rings plus scattered noise: shapes no centroid can capture.
str(rings)
```

## What every result has in common

```{r shared}
fit <- shoal_kmeans(x, k = 3L)

fit$cluster[1:10]
fit$n_clusters
fit$n_noise
fit$params
```

`data` holds the matrix the model was fitted to, so `plot()` needs nothing
else. With two columns it draws a scatter plot; with more, a scatter plot
matrix, or a chosen pair via `xcol` and `ycol`.

```{r shared-plot}
plot(fit, xcol = "Petal.Length", ycol = "Petal.Width")
```

## k-means: `shoal_kmeans()`

Partitions the data into `k` spherical clusters of similar size by minimising
the within-cluster sum of squares. It is fast, predictable, and the right
baseline when clusters are compact and you know roughly how many to expect.
It is stochastic in its initialisation; `seed` controls that, and R's own
random number generator has no effect on it.

```{r kmeans}
km <- shoal_kmeans(x, k = 3L, n_runs = 10L, seed = 1L)
km
km$centroids
```

Because it has cluster centres, it can assign new observations.

```{r kmeans-predict}
predict(km, x[c(1, 51, 101), ])
```

`inertia` decreases with `k` by construction, so it locates a point of
diminishing returns rather than an optimum. See [Choosing the number of
clusters](#choosing-k) for measures that do have an interior optimum.

## Gaussian mixture: `shoal_gmm()`

Fits `k` multivariate Gaussians by expectation-maximisation. Clusters are
elliptical rather than spherical, so correlated or differently scaled features
are handled without pre-scaling, and every observation gets a probability of
belonging to each component rather than a hard label. `cluster` is the most
likely component; `posterior` has the full matrix.

```{r gmm}
gm <- shoal_gmm(x, k = 3L)
gm

round(gm$posterior[c(1, 51, 101), ], 3)
```

A fitted mixture has a `logLik()` method, so `AIC()` and `BIC()` work on it
directly. Only full covariance matrices are supported.

```{r gmm-bic}
BIC(gm)
```

## DBSCAN: `shoal_dbscan()`

Grows clusters from points that have at least `min_samples` neighbours within
radius `eps`, and labels everything else noise. Clusters can be any shape, but
they must share one density, because `eps` is a single global threshold. The
rings are an easy case.

```{r dbscan}
db <- shoal_dbscan(rings, eps = 0.2, min_samples = 5L)
db
plot(db)
```

`eps` is on the scale of the data, so the same value will not carry between
datasets. `metric = "cosine"` is available for directional data.

## HDBSCAN: `shoal_hdbscan()`

Extends DBSCAN across every density at once by building a hierarchy of
clusterings and keeping the most stable ones. It has no `eps`; instead
`min_cluster_size` sets the smallest group worth calling a cluster and
`min_samples` sets how conservative the density estimate is. Larger values of
either give fewer, larger clusters and more noise.

```{r hdbscan}
hdb <- shoal_hdbscan(rings, min_cluster_size = 15L, min_samples = 5L)
hdb
```

It also returns a GLOSH outlier score for every observation, near 1 for points
far from any cluster.

```{r hdbscan-glosh}
head(order(hdb$outlier_scores, decreasing = TRUE))
```

Both density algorithms rely on a spatial index that degrades as the number of
columns grows. For wide data, either reduce the dimension first (see the UMAP
vignette) or, for embedding vectors, use EVoC.

## Hierarchical clustering: `shoal_hclust()`

Merges observations bottom-up into a dendrogram, which can then be cut at any
number of clusters. It takes a `dist` object, which `shoal_dist()` computes in
Rust, or raw data. The result is a standard `hclust`, so `cutree()`,
`plot()` and `as.dendrogram()` all apply.

```{r hclust}
d <- shoal_dist(x)
hc <- shoal_hclust(d, method = "ward")

table(cutree(hc, k = 3), iris$Species)
plot(hc, labels = FALSE, hang = -1)
```

Seven linkage methods are available. Note that `"ward"` is R's `"ward.D2"`,
and that `"centroid"` and `"median"` take plain distances where
`stats::hclust()` expects squared ones; see `?shoal_hclust`.

## EVoC: `shoal_evoc()`

For embedding vectors: rows that are directions in a high-dimensional space,
such as the output of a text or image model. EVoC builds a nearest-neighbour
graph under cosine distance, learns a compact node embedding from it, and
density-clusters that embedding at several granularities at once. All layers
are returned; `layer` picks which one populates `cluster`, by default the
most persistent.

```{r evoc}
set.seed(1)
centres <- matrix(runif(6 * 48, -1, 1) * 0.6, nrow = 6)
emb <- centres[rep(1:6, each = 140L), ] +
  matrix(rnorm(840 * 48, sd = 0.1), ncol = 48)

ev <- shoal_evoc(emb, min_cluster_size = 15L)
ev
```

EVoC is the wrong tool for ordinary tabular data: it assumes cosine geometry
and is calibrated for thousands to millions of rows. The upstream default
`min_cluster_size = 5` over-fragments small collections, so raise it first.

## Choosing the number of clusters {#choosing-k}

`shoal_metrics()` computes the Calinski-Harabasz index (higher is better) and
Davies-Bouldin index (lower is better) for any partition, from the data.
`shoal_silhouette()` scores every observation from a distance matrix, and
works for any cluster shape.

```{r choose-k}
ks <- 2:6
indices <- do.call(rbind, lapply(ks, function(k) {
  shoal_metrics(shoal_kmeans(x, k = k))
}))
indices$bic <- vapply(ks, function(k) BIC(shoal_gmm(x, k = k)), numeric(1))
indices

sil <- shoal_silhouette(d, km)
attr(sil, "avg_width")
```

The indices disagree on this data, and that is the right answer: two of the
species overlap, so two and three clusters are both defensible. shoal reports
the evidence and leaves the decision to you.

## Which algorithm?

| You have                                     | Start with                        |
|----------------------------------------------|-----------------------------------|
| Compact clusters, a known `k`                | `shoal_kmeans()`                  |
| Elliptical clusters, or you want probabilities | `shoal_gmm()`                   |
| Arbitrary shapes at one density              | `shoal_dbscan()`                  |
| Arbitrary shapes at varying density          | `shoal_hdbscan()`                 |
| A dendrogram, or cuts at several `k`         | `shoal_hclust()`                  |
| Embedding vectors, thousands of rows or more | `shoal_evoc()`                    |
