---
title: "Clustering after UMAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering after UMAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

Clustering wide data directly is hard for every algorithm here, and hardest
for the density-based ones: distances concentrate as dimension grows, so the
contrast between dense and sparse regions that DBSCAN and HDBSCAN depend on
fades, and the spatial index behind them slows down. A common remedy is to
embed the data into a few dimensions with UMAP first, then cluster the
embedding. This vignette shows what that does to each algorithm on one real
dataset, and where it helps, where it does not, and what to watch for.

UMAP comes from the [uwot](https://cran.r-project.org/package=uwot) package;
the data from [mlbench](https://cran.r-project.org/package=mlbench). Neither
is required by shoal.

```{r setup, message = FALSE}
library(shoal)
library(uwot)
library(mlbench)
```

## The data

`Satellite` holds 6,435 pixels of Landsat imagery, each described by 36
spectral values, labelled with one of six land-cover classes. Three of the
classes are grades of the same thing: grey soil, damp grey soil and very damp
grey soil.

```{r data}
data(Satellite)
x <- scale(as.matrix(Satellite[, 1:36]))
truth <- Satellite$classes
table(truth)
```

The labels are used only to score results, with the adjusted Rand index (ARI):
1 for a perfect match, 0 for chance agreement. Noise points count as their own
class, so an algorithm that discards data pays for it.

```{r ari}
ari <- function(a, b) {
  a <- ifelse(is.na(a), -1L, as.integer(a))
  b <- ifelse(is.na(b), -1L, as.integer(b))
  tab <- table(a, b)
  comb2 <- function(x) x * (x - 1) / 2
  sum_ij <- sum(comb2(tab))
  sum_a <- sum(comb2(rowSums(tab)))
  sum_b <- sum(comb2(colSums(tab)))
  expected <- sum_a * sum_b / comb2(sum(tab))
  (sum_ij - expected) / ((sum_a + sum_b) / 2 - expected)
}

# Every algorithm on one input, as a one-row summary each.
run_all <- function(data, eps, min_cluster_size) {
  fits <- list(
    "k-means"  = shoal_kmeans(data, k = 6L),
    "GMM"      = shoal_gmm(data, k = 6L),
    "DBSCAN"   = shoal_dbscan(data, eps = eps, min_samples = 10L),
    "HDBSCAN"  = shoal_hdbscan(data, min_cluster_size = min_cluster_size,
                               min_samples = 10L, boruvka = ncol(data) <= 10L)
  )
  hc <- shoal_hclust(shoal_dist(data), method = "ward")
  rows <- lapply(names(fits), function(nm) {
    f <- fits[[nm]]
    data.frame(algorithm = nm, clusters = f$n_clusters, noise = f$n_noise,
               ari = round(ari(truth, f$cluster), 3))
  })
  rows[["hclust"]] <- data.frame(algorithm = "Ward (cut at 6)", clusters = 6L,
                                 noise = 0L, ari = round(ari(truth, cutree(hc, 6)), 3))
  out <- do.call(rbind, rows)
  rownames(out) <- NULL
  out
}
```

## Clustering the raw 36 dimensions

```{r raw}
raw <- run_all(x, eps = 2, min_cluster_size = 100L)
raw
```

The centroid and hierarchical methods find something, because the classes do
differ in their mean spectra. The density methods find almost nothing: two
clusters and hundreds of noise points, an ARI near zero. It is not a parameter
problem. In 36 standardised dimensions the nearest-neighbour distances of
almost every point are similar, so there is no density contrast to work with,
and no `eps` or `min_cluster_size` recovers it.

## Embedding with UMAP

UMAP builds a nearest-neighbour graph in the original space and lays it out
in a few dimensions so that neighbours stay close. Two settings matter when
the goal is clustering rather than a picture:

- `min_dist = 0` lets points pack tightly, which sharpens the density
  contrast the density algorithms need. The visual default of 0.1 spreads
  clusters out for legibility.
- `n_neighbors` trades local detail for global structure. Values around 30
  give clusters that are stable across runs; 15 fragments more.

UMAP is stochastic, so the seed is set, and the result is checked against
other seeds below.

```{r umap}
set.seed(42)
u <- umap(x, n_neighbors = 30, min_dist = 0, n_threads = 2)
colnames(u) <- c("UMAP1", "UMAP2")

pal <- grDevices::hcl.colors(6, "Dark 3")
plot(u, col = pal[as.integer(truth)], pch = 19, cex = 0.4,
     main = "UMAP embedding, coloured by class")
legend("topleft", legend = levels(truth), col = pal, pch = 19, bty = "n", cex = 0.8)
```

Red soil, cotton crop and vegetation stubble each get an island of their own.
The three grey-soil classes form one continuous band, shading from grey soil
at one end to very damp grey soil at the other. That is the shape of the data,
and it decides what each algorithm can do next.

## Clustering the embedding

`eps` for DBSCAN is on the scale of the embedding, not the original data, so
it is chosen afresh.

```{r embedded}
embedded <- run_all(u, eps = 0.3, min_cluster_size = 100L)

comparison <- merge(raw, embedded, by = "algorithm", suffixes = c("_raw", "_umap"))
comparison[order(-comparison$ari_umap), c("algorithm", "ari_raw", "ari_umap",
                                          "clusters_umap", "noise_umap")]
```

Three different things happened.

**The density methods went from useless to sound.** DBSCAN and HDBSCAN now
find the islands with almost no noise, and any extra clusters they report are
small fragments at the edges. Their ARI is capped near 0.46 because the
grey-soil band comes out as one cluster: it is one dense region, with no
density gap at which to split it into three grades. That is a limit of the
labels' relationship to the data, not a failure to see the structure.

```{r hdbscan-table}
hdb <- shoal_hdbscan(u, min_cluster_size = 100L, min_samples = 10L)
table(cluster = hdb$cluster, truth, useNA = "ifany")
plot(hdb)
```

**The model-based methods improved because their assumptions became true.**
In 36 dimensions the Gaussian mixture was fitting ellipsoids to classes that
are not ellipsoidal; on the embedding, each island is roughly one blob, and
the six-component mixture cuts the band into thirds along its length. Ward
linkage gains the most for the same reason. Both are rewarded by the labels
for slicing a gradient into the pieces the labels happen to use.

**k-means barely changed.** It was already finding the class means in the
original space, and a six-way partition of the embedding is no better than a
six-way partition of the raw data. UMAP helps algorithms that need local
structure; it does little for one that only needs centroids.

## Stability across UMAP seeds

A UMAP layout is one draw from a stochastic optimisation. Whether the
clusters found on it are real is a question of whether another draw gives the
same clusters. Two more seeds:

```{r seeds}
seed_results <- lapply(c(1, 2), function(s) {
  set.seed(s)
  us <- umap(x, n_neighbors = 30, min_dist = 0, n_threads = 2)
  list(
    hdbscan = shoal_hdbscan(us, min_cluster_size = 100L, min_samples = 10L)$cluster,
    ward = cutree(shoal_hclust(shoal_dist(us), method = "ward"), 6)
  )
})

data.frame(
  method = c("HDBSCAN", "Ward (cut at 6)"),
  seed1_vs_seed2 = c(
    round(ari(seed_results[[1]]$hdbscan, seed_results[[2]]$hdbscan), 3),
    round(ari(seed_results[[1]]$ward, seed_results[[2]]$ward), 3)
  ),
  seed1_vs_truth = c(
    round(ari(truth, seed_results[[1]]$hdbscan), 3),
    round(ari(truth, seed_results[[1]]$ward), 3)
  ),
  seed2_vs_truth = c(
    round(ari(truth, seed_results[[2]]$hdbscan), 3),
    round(ari(truth, seed_results[[2]]$ward), 3)
  )
)
```

HDBSCAN returns nearly the same partition whichever layout it is given,
because the islands are the same; only the number of small fragments varies.
Ward's cut moves with the layout, because where a continuous band gets
divided into three depends on the exact shape it was drawn in. Its higher
score against the labels is partly the luck of the draw. When a clustering is
going to be acted on, the stable one is worth more than the one that happened
to score best.

## Where EVoC fits

`shoal_evoc()` does something similar internally: a nearest-neighbour graph,
a learned node embedding of it, and a density clustering of that at several
granularities, in one call. The difference is that it assumes cosine
geometry, which is right for embedding vectors from a text or image model and
wrong for spectral measurements like these. On this data it finds a few
clusters and labels a large share of the input as noise; on the data it is
built for, it replaces the whole UMAP-then-HDBSCAN recipe.

```{r evoc}
ev <- shoal_evoc(x, min_cluster_size = 100L)
ev$n_clusters
ev$n_noise
```

## Practical notes

- Standardise the columns before UMAP unless they are already on one scale;
  the neighbour graph is built from Euclidean distances in the original space.
- Choose `eps` on the embedding, not the data. Distances in a UMAP layout are
  not the distances you started with, and are not comparable between runs.
- Two dimensions are enough for the algorithms here. Embedding to five gave
  the same clusters on this data, at the cost of not being able to plot them.
- Report the seed, and check another. A clustering that survives a second
  layout is structure; one that does not is an artefact of the drawing.
- Do not use `shoal_metrics()` or `shoal_silhouette()` on the embedding to
  compare against raw-space results: the scores live in different spaces.
  They remain useful for comparing clusterings of the same embedding.
